---
title: "Smoking Paper - Appendix A" 
subtitle: "Genus level"
author: "Giacomo Antonello"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    self_contained: true
    code_folding: hide
    toc_depth: 2
    toc_float: true
    number_sections: false
    thumbnails: false
    lightbox: true
    gallery: false
    use_bookdown: true
    highlight: haddock
---

```{r, echo=FALSE}

knitr::opts_chunk$set(
  cache = FALSE,
  concordance = TRUE,
  prompt = TRUE, # fig.width=5, fig.height=5,
  out.width = "100%",
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  error = TRUE,
  tidy = FALSE,
  comment = ""
  )

```

# Data preparation {.tabset}

```{r library setup}
# from Bioconductor (install by running: BiocManager::install("x") | x = name of the package)
# BiocManager::install(c("DESeq2", "phyloseq", "microbiome"))

#github functions
# devtools::install_github("bryandmartin/corncob")
# remotes::install_github("g-antonello/gautils")
# remotes::install_github("mikemc/speedyseq")#
# devtools::install_github("gaospecial/ggvenn")

library(gautils)
library(magrittr) # imported separately from tidyverse to also import '%<>%'
library(ggpubr)
library(kableExtra)

# graphics packages
# library(ggpubr)
# library(magrittr)
# # my functions
# source("my_personal_functions.R")
# 
# 
# library(speedyseq)
theme_set(theme_light())


chrismb_phy <- readRDS("/shared/statgen/microbiome/CHRISMB/processed_data/microbiome_tables/phyloseq_built.Rds")


threshold_prevalence <- 0.01
threshold_detection <- 10


```

## Download CHRIS data

These are based on data retrieval from the `chrisData` internal R package curated by J. Rainer. Variable selection is done via consulting the codebook. 

```{r, message=FALSE, warning=FALSE}
chrismb_phy <- readRDS("/shared/statgen/microbiome/CHRISMB/processed_data/microbiome_tables/phyloseq_built.Rds")

library(chrisData)

possible_data_files <- chrisDataFile()

selected_data_files <- c(
    "general_information",
    "household",
    "clinical_traits",
    "interview",
    "labdata",
    "substudies",
    "drugs_summary"
  )

data_selected.list <- sapply(selected_data_files, function(x) chrisData(paste0(x, ".txt.gz")), USE.NAMES = T)

data_selected.joined.df <- purrr::reduce(data_selected.list, full_join, by = "AID")
data_selected.joined.df <- data_selected.joined.df[match(chrismb_phy@sam_data$AID, data_selected.joined.df$AID),]

# Columns with all NAs must be necessarily removed 
all_NAs <- apply(data_selected.joined.df, 2, function(x) all(is.na(x)))
data_selected.joined.df <- data_selected.joined.df[, !all_NAs]

# Columns of class "chron" are problematic, they must be removed to be able to work with phyloseq and metadata data.frame back and forth
data_selected.joined_NODates.df <- select(data_selected.joined.df, -contains(c("x0bp08", "x0bc03", "x0bc06")))

rm(data_selected.list)

chrismb_phy <- phy_add_metadata_variables(chrismb_phy, df = data_selected.joined_NODates.df, by = "AID")


levels(chrismb_phy@sam_data$x0oh01) <- c("0", "1-9", "10-19", "20+")
levels(chrismb_phy@sam_data$x0sm51) <- c("Never", "Former", "Current (R)", "Current (NR)")
chrismb_phy@sam_data$x0an03a %<>% as.numeric()

chrismb_phy@sam_data$x0_ager_cat <- cut(chrismb_phy@sam_data$x0_ager, breaks = c(18, 31, 41, 51, 61, 71, max(chrismb_phy@sam_data$x0_ager)+1), include.lowest = TRUE, labels = c("18-30", "31-40", "41-50", "51-60", "61-70", "71+"))

# create some househod (hid) info
chrismb_phy <- relocate_sample_data(chrismb_phy, x0_ager_cat, .after = x0_ager) %>% 
  mutate_sample_data(
    hid = paste0("h", hid),
    hid_size_chrismb = add_N_to_catVar(hid) %>% as.character() %>% gsub("h.*\\ ", "", .) %>% parse_number(),
  smoking_exposure_ga = 
    case_when(
      grepl("Current", as.character(x0sm51)) ~ "Current",
      TRUE ~ as.character(x0sm51)
      ) %>% 
    factor(levels = c("Never", "Former", "Current"))
  ) %>% 
  relocate_sample_data(smoking_exposure_ga, .before = x0sm51)

```

## Smoking variables generation

### Bin smoking intensity (g/day)

```{r}

###############################
## bin daily tobacco smoked
###############################

chrismb_phy %<>%
  mutate_sample_data(
    ### tobacco g per day (bin 5 by 5)
    x0sm61_bin5cat = case_when(smoking_exposure_ga == "Current" & !is.na(x0sm61) ~ cut(
      x0sm61,
      breaks = c(0, 1, 2, seq(5, 30, 5), 61),
      include.lowest = TRUE
    )),
  ### tobacco g per day (bin 5 by 5, semi-continuous, using upper limit of the range as integer)
  x0sm61_bin5cont = case_when(smoking_exposure_ga == "Current" & !is.na(x0sm61) ~ cut(
      x0sm61,
      breaks = c(0, 1, 2, seq(5, 30, 5), 60),
      labels = c(1, 2, seq(5, 30, 5), 60),
      include.lowest = TRUE
    )) %>% as.character() %>% as.numeric(),
  
    ### tobacco g per day (bin 2 by 2)
    x0sm61_bin2cat = case_when(smoking_exposure_ga == "Current" & !is.na(x0sm61) ~ cut(
      x0sm61,
      breaks = c(0, 1, seq(2, 30, 2), 60),
      include.lowest = TRUE
    )),
  
  ### tobacco g per day (bin 2 by 2, semi-continuous, using upper limit of the range as integer)
  x0sm61_bin2cont = case_when(smoking_exposure_ga == "Current" & !is.na(x0sm61) ~ cut(
      x0sm61,
      breaks = c(0, 1, seq(2, 30, 2), 60),
      labels = c(1, seq(2, 30, 2), 60),
      include.lowest = TRUE
    )
    ) %>% as.character() %>% as.numeric()
  )

```

### Bin Years since quitting

```{r}
# x0sm40 = age when people quit or reduce
# x0_ager = age rounded to closest integer

max_value <- max(chrismb_phy@sam_data$x0_ager - chrismb_phy@sam_data$x0sm40, na.rm = T)

chrismb_phy %<>% 
  mutate_sample_data(
    years_since_quitting_or_reducing = x0_ager - x0sm40,
    
    years_since_quitting_binned = case_when(
      smoking_exposure_ga == "Former" & !is.na(years_since_quitting_or_reducing) ~ cut(years_since_quitting_or_reducing, 
                                            breaks = c(0,1,3,5,10,max_value), 
      include.lowest = TRUE
    )
  )
)

rm(max_value)

```

### Ultimate smoking variable for plots: "smoking_detailed"

```{r}

chrismb_phy %<>% 
  mutate_sample_data(
    smoking_detailed = case_when(
      smoking_exposure_ga == "Never" ~ "Never",
      
      smoking_exposure_ga == "Former" ~ paste(smoking_exposure_ga,
                                               years_since_quitting_binned, 
                                               "y"),
      smoking_exposure_ga == "Current" ~ paste(smoking_exposure_ga,
                                               x0sm61_bin5cat,
                                               "g"),
      TRUE ~ "Other"
    ),
    smoking_detailed.Extra = case_when(
      smoking_exposure_ga == "Never" ~ "Never",
      
      smoking_exposure_ga == "Former" ~ paste(smoking_exposure_ga,
                                               years_since_quitting_or_reducing, 
                                               "y"),
      smoking_exposure_ga == "Current" ~ paste(smoking_exposure_ga,
                                               x0sm61,
                                               "g"),
      TRUE ~ "Other"
    )
    
  )



nice_order_current <- sort(grep("Current", unique(chrismb_phy@sam_data$smoking_detailed), value = TRUE))[c(9,1,4,8,2,3,5,6,7,10)] %>% rev()

nice_order_former <- sort(grep("Former", unique(chrismb_phy@sam_data$smoking_detailed), value = TRUE))[c(5,1,3,4,2,6)]

final_levels <- c(nice_order_current,nice_order_former, "Never")


# SMOKING DETAILED 1 - make ordered factor, useful for plotting
chrismb_phy@sam_data$smoking_detailed <- factor(chrismb_phy@sam_data$smoking_detailed, levels = final_levels)


```

## Available Samples for statistical analysis

```{r}
excluded_samples_upon_request <- readLines("/home/gantonello/CHRISMB/participants lists each study/withdrawnConsents_AIDs.txt")

chrismb_phy_core <- core(chrismb_phy, detection = threshold_detection, prevalence = threshold_prevalence)

phy_Q1 <- chrismb_phy_core %>% 
  filter_sample_data(!(AID %in% excluded_samples_upon_request),
                     !is.na(x0oh01),
                     !is.na(x0_ager),
                     !is.na(x0_sex),
                     !is.na(x0sm51),
                     #x0bc10 != 1, # people who smoked were reported as 1, NAs are treated as non-smoking
                     x0dd51 == "No") %>% 
  core(detection = 10, prevalence = 1/100)

phy_Q1


phy_Q1_genus <- phy_tax_glom(phy_Q1, "Genus")


phy_Q1_meta <- meta(phy_Q1)

```

# Table 1 - Diastolic Blood pressure is significantly lower in smokers {.tabset}

## Overall 

```{r}

bp_model0 <- lm(x0bp02 ~ x0_sex + x0_ager + x0an03a + x0oh01 + x0sm51, data = phy_Q1_meta)

broom::tidy(bp_model0) %>% 
  filter(p.value < 0.05) %>% 
  mutate(p.value = format(p.value, scientific = TRUE, digits = 3)) %>%
  mutate_if(is.numeric, round, 2) %>% 
  kbl(caption = "Blood pressure is significantly lower in current smokers, after adjusting for age, sex, body mass index (BMI) and the number of teeth. We included the number of teeth in the model to account for potential hypertensive effect of poor oral health, in the context of the literature regarding blood pressure, edentulism and nitrix oxide microbial conversion in the oral cavity") %>% 
  kable_styling()

```

## 18 - 45 years

```{r}

bp_model0 <- lm(x0bp02 ~ x0_sex + x0an03a + x0sm51, data = phy_Q1_meta %>% filter(age <= 45))

broom::tidy(bp_model0) %>% 
  filter(p.value < 0.05) %>% 
  mutate(p.value = format(p.value, scientific = TRUE, digits = 3)) %>%
  mutate_if(is.numeric, round, 2) %>% 
  kbl(caption = "Blood pressure is significantly lower in current smokers younger than 45 years, after adjusting for age, sex, body mass index (BMI) and the number of teeth. We included the number of teeth in the model to account for potential hypertensive effect of poor oral health, in the context of the literature regarding blood pressure, edentulism and nitrix oxide microbial conversion in the oral cavity") %>% 
  kable_styling()

```

## 46 - `r max(phy_Q1_meta$x0_ager, na.rm = T)` years

```{r}

bp_model0 <- lm(x0bp01 ~ x0_sex + x0an03a + + x0_ager + x0sm51, data = phy_Q1_meta %>% filter(age > 45))

broom::tidy(bp_model0) %>% 
  filter(p.value < 0.05) %>% 
  mutate(p.value = format(p.value, scientific = TRUE, digits = 3)) %>%
  mutate_if(is.numeric, round, 2) %>% 
  kbl(caption = "Blood pressure is significantly lower in current smokers older than 45 years, after adjusting for age, sex, body mass index (BMI) and the number of teeth. We included the number of teeth in the model to account for potential hypertensive effect of poor oral health, in the context of the literature regarding blood pressure, edentulism and nitrix oxide microbial conversion in the oral cavity") %>% 
  kable_styling()

```


# Table 2 - Body Mass Index is significantly higher in smokers


## Overall 

```{r}

bmi_model0 <- lm(x0an03a ~ x0_ager + x0oh01 + x0_sex + x0bp02 + x0sm51, data = phy_Q1_meta)

broom::tidy(bmi_model0) %>% 
  filter(p.value < 0.05) %>% 
  mutate(p.value = format(p.value, scientific = TRUE, digits = 3)) %>% 
  mutate_if(is.numeric, round, 2) %>% 
  kbl(caption = "Body Mass Index (BMI) is significantly higher in current smokers, after adjusting for age, sex, systolic blood pressure and the number of teeth") %>% 
  kable_styling()
  
```

## 18 - 45 years

```{r}

bmi_model0 <- lm(x0an03a ~ x0oh01 + x0_sex + x0bp02 + x0sm51, data = phy_Q1_meta %>% filter(age <= 45))

broom::tidy(bmi_model0) %>% 
  filter(p.value < 0.05) %>% 
  mutate(p.value = format(p.value, scientific = TRUE, digits = 3)) %>% 
  mutate_if(is.numeric, round, 2) %>% 
  kbl(caption = "Body Mass Index (BMI) is significantly higher in current smokers, after adjusting for age, sex, systolic blood pressure and the number of teeth") %>% 
  kable_styling()
  
```


## 46 - `r max(phy_Q1_meta$age, na.rm = T)` years

```{r}

bmi_model0 <- lm(x0an03a ~ x0oh01 + x0_sex + x0bp02 + x0sm51, data = phy_Q1_meta %>% filter(age > 45))

broom::tidy(bmi_model0) %>% 
  filter(p.value < 0.05) %>% 
  mutate(p.value = format(p.value, scientific = TRUE, digits = 3)) %>% 
  mutate_if(is.numeric, round, 2) %>% 
  kbl(caption = "Body Mass Index (BMI) is significantly higher in current smokers, after adjusting for age, sex, systolic blood pressure and the number of teeth") %>% 
  kable_styling()

```

